Last glacial cycle glacier erosion potential in the Alps

### 2021-02-09 articles/erosion/alps-erosion.tex
Add refs to Olivia's papers, Antarctic speed-up.

    That said, the magnitude of the modelled erosion rates can hardly be used
    as a criteria to rule out this or that erosion law for the glacial Alps.
    Firstly, while sliding is here modelled by a pseudo-plastic law based on
    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
    empirical erosion rules are thought to be a proxy for glacier abrasion
    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
    over a hard bed. Secondly, it is
    arguable that a layer of till and proglacial sediments may have armored the
    glacier bed from erosion for at least parts of the glacial cycle. Given the
    lack of data to constrain the erosion rule, it can not
    be excluded that a linear or sub-linear erosion-law
    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
    Even then, the intrinsically large variations in modelled glacier sliding
    velocities yield cumulative glacial-cycle erosion two to three orders of
    magnitude higher in the Alpine valleys than near the mountain tops
    (Fig.~\ref{fig:powerlaws}d).

@@

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \citep[cf.][for comparison]
    {Werder.etal.2013, Iverson.2012, Ugelvig.etal.2018}. The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][for the physics]
    {Aschwanden.etal.2012, Seguinot.etal.2016}. Observed feedbacks between
    surface meltwater and basal sliding, not included in the model, presumably
    further accelerated erosion during deglaciation \citep{Herman.etal.2011}.
    Furthermore, we omit the fluvial transport of sediment, which could well
    increase sediment discharge and erosion with increased melt, so long as
    additional sediment remains available \citep{Delaney.Adhikari.2020}. Rapid
    glacier erosion can be expected to accompany the ongoing 21st-century
    deglaciation, especially for the ice sheets of Greenland and
    Antarctica where polythermal and cold-based glaciers are warming and
    sliding may increase \citep[e.g.,][]{Moon.etal.2012, Overeem.etal.2017,
    Mouginot.etal.2019}.

>>>>

    That said, the magnitude of the modelled erosion rates can hardly be used
    as a criteria to rule out this or that erosion law for the glacial Alps.
    While sliding is here modelled by a pseudo-plastic law based on
    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
    empirical erosion rules are thought to be a proxy for glacier abrasion
    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
    over a hard bed. Current erosion rates in the Alps strongly vary depending
    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
    it arguable that a layer of till and proglacial sediments may have armored
    the glacier bed from erosion for at least parts of the glacial cycle.
    Given the lack of data to constrain the erosion rule, it can not
    be excluded that a linear or sub-linear erosion-law
    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
    Even then, the intrinsically large variations in modelled glacier sliding
    velocities yield cumulative glacial-cycle erosion two to three orders of
    magnitude higher in the Alpine valleys than near the mountain tops
    (Fig.~\ref{fig:powerlaws}d).

@@

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \citep[cf.][for comparison]
    {Werder.etal.2013, Iverson.2012, Ugelvig.etal.2018}. The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][for the physics]
    {Aschwanden.etal.2012, Seguinot.etal.2016}. Observed feedbacks between
    surface meltwater and basal sliding, not included in the model, presumably
    further accelerated erosion during deglaciation \citep{Herman.etal.2011}.
    Furthermore, we omit the fluvial transport of sediment, which could well
    increase sediment discharge and erosion with increased melt, so long as
    additional sediment remains available \citep{Delaney.Adhikari.2020}. Rapid
    glacier erosion can be expected to accompany the ongoing 21st-century
    deglaciation, especially for the ice sheets of Greenland and
    Antarctica where polythermal and cold-based glaciers are warming and
    sliding may increase \citep[e.g.,][]{Moon.etal.2012, Mouginot.etal.2014,
    Overeem.etal.2017}.
